---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---




```{r warning=FALSE, echo=FALSE}

###################################################################
# Load Libraries 
###################################################################
library(tidyverse)

###################################################################
# Load Tables that are stored as CSVs on Github
###################################################################

# ICD-9 Diagnostic codes
nineDX11 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/9DX11.csv")

# ICD-9-CM Operative Codes
nine0P11 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/9OP11.csv?token=AOUMBHAGKGSZG2TSGOG2CP3AM2FTQ")

# Patient ICU stays
CRIT_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/CRIT_2000.csv?token=AOUMBHADQBHKXUUKB7V3Q4TAM2GDY") %>%
  select(-PAT_NO_TAG)

# Disposition codes
DISPOSTN <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/DISPOSTN.csv?token=AOUMBHE4FATYKQPJTDSMMF3AM2GE4")

# DRGs
DRG <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/DRG.csv?token=AOUMBHFBTBYE3Q4TL4KTRRDAM2GF4")

# Inpatient encounters
OMEGA_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/OMEGA_2000.csv?token=AOUMBHA2FQ7AWMLVZE5FCQ3AM2GHG")

# Patient diagnostic data
OME_DX_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/OME_DX_2000.csv?token=AOUMBHFOYLTZQZQBGMTEAILAM2GIU")

# Patient operative data
OME_PX_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/OME_PX_2000.csv?token=AOUMBHA66FZEP7AHOSTBOKDAM2GKY")

# Organ donations
ORGAN <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/ORGAN.csv?token=AOUMBHFMVUHRBMG6KI3IDRDAM2GMC")

# ???????? This table is provided in the database but is not described in the codebook
PAYEX <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/PAYEX.csv?token=AOUMBHDRDOXPNZORSPDOJSLAM2GNC")

# Services codes
SERVICE <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/SERVICE.csv?token=AOUMBHBANABWFQP7P3GVNALAM2GOC")

# Admission Source
SOURCE <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/SOURCE.csv?token=AOUMBHHIEF5IJH3KPSV45STAM2GPQ")

# Tissues
TISSUE <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/TISSUE.csv?token=AOUMBHFMOZUROW2XOBCCVQTAM2GQM")

# Yes/No designation
YES_NO <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/YES_NO.csv?token=AOUMBHA2GSK3P6IFX4HAYJDAM2GTC")

```

# Examining Referential Integrity

## CRIT_2000 | Patient data table regarding ICU stays

No primary key. Is PAT_NO-ADM_DATE the composite key?
```{r}
CRIT_2000 %>%
  count(PAT_NO, ADM_DATE) %>%
  arrange(desc(n)) %>%
  slice(1)
```
Nope, a single PAT_NO-ADM_DATE combo appears six times.

Let's try adding in ICU.
```{r}
CRIT_2000 %>% count(PAT_NO, ADM_DATE, ICU) %>%
  arrange(desc(n))
```
Nope, didn't work either. Let's try PAT_NO-ADM_DATE-OCCURRENCE

```{r}
CRIT_2000 %>% count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n)) %>%
  slice(1)
```
That worked. Our **composite key is PAT_NO-ADM_DATE-OCCURRENCE**.

Each row is a unique combination of patient id, date of admission, and occurrence. The definition of occurrence in the codebook is confusing: "Ordering of CCU Stays." What is a Critical Care Unit, and how is it difference from the ICU? And how does it work out that someone ivsits the CCU multiple separate times over their admission?


And what's the difference between all the ICU categories?
```{r}
CRIT_2000 %>% count(ICU)
```
Google says MICU is "medical intensive care unit". SICU is "surgical intensive care unit". I can't seem to figure out what CCU and SPEC_NSRY stand for.


What's the difference between ICU_LOS and TOT_CCUS? Is "critical care" different from "intensive care"? Both lengths of stay added together are usually far less than the total stay (discharge - admitted). 


## DRG

What's the primary key in the DRG table?
```{r}
DRG %>% count(DRG) %>%
  arrange(desc(n)) %>%
  slice(1)
```
It looks like **DRG is the DRG Primary key**.

What is the WEIGHT and ARITH_LOS and MEAN_LOW and CUTOFF and how is reimbursement calculated? Are we supposed to use these in calculating reimbursement by DRG?. 


## nineDX11

What is the primary key in nineDX11?
```{r}
nineDX11 %>% count(DIAG) %>%
  arrange(desc(n)) %>%
  slice(1)
```
It appears there can be up to two rows for each DIAG in the table. Why?
```{r}
nineDX11 %>%
  group_by(DIAG) %>%
  mutate(totes_diags = n()) %>%
  ungroup() %>%
  filter(totes_diags > 1) %>%
  arrange(DIAG)
```
It seems the rows are complete duplicates, but let's check that's true if we do a group_by for all columns. 

```{r}
nineDX11 %>%
  group_by(DIAG) %>%
  mutate(totes_diags = n()) %>%
  ungroup() %>%
  filter(totes_diags > 1) %>%
  group_by(DIAG, DIAG_DESC, DIAG_C, DIAG_5) %>%
  count() %>%
  arrange()
```
It seems indeed these rows are completely duplicated, because none of them had differences across any of the variables. 

Safe to filter out the duplicate values I think?

```{r}
nineDX11 <- nineDX11 %>%
  distinct()
```
**DIAG is the primary key.**

## nineOP11 | Descriptor Table for ICD-9-CM Operative Codes 

What's the primary key for nineOP11?

```{r}
nine0P11 %>%
  count(OP) %>%
  arrange(desc(n)) %>%
  slice(1)
```
Seems like **OP is the primary key for nineOP11**. All fine here. Could probably use OP_C as well.


## DISPOSTN

Is DISPOSTN the primary key of DISPOSTN?
```{r}
DISPOSTN %>% count(DISPOSTN) %>%
  arrange(desc(n)) %>%
  slice(1)
```
Yep, all fine here. 


## PAYEX

There's no table description for PAYEX in the documentation. Does that mean it's safe to assume we don't need to use it?

It's a small table. Don't need to check counts. 


## ORGAN

Small table, looks fine. **ORGAN_DON** is the primary key.


## SERVICE

Small table, looks fine. **SERVICE** is the primary key.

## SOURCE 

Small table, looks fine. **ADM_SOURCE** is the primary key.


## TISSUE

Small table, looks fine. **TISSUE** is the primary key.


## YES_NO

Small table, looks fine. **YES_NO** is the primary key. But where does this link to other tables?


## OMEGA_2000 | Data for each inpatient encounter during 2000

What's the primary key in the OMEGA_2000 table?

Start with PAT_NO & ADM_DATE & ADM_HR
```{r}
OMEGA_2000 %>% count(PAT_NO, ADM_DATE, ADM_HR) %>%
  arrange(desc(n)) %>% 
  slice(1) 
```
Nope, we have duplicates of that pairing.

Are there genuinely duplicated rows in the table?

```{r}
OMEGA_2000 %>%
  summarize(n())
OMEGA_2000 %>%
  distinct() %>%
  summarize(n())
```

Yes, it seems there are. Let's remove duplicates.
```{r}
OMEGA_2000 <- OMEGA_2000 %>%
  distinct()
```

Let's check if PAT_NO-ADM_DATE-ADM_HR is now a primary key now that overall duplicates removed 

```{r}
OMEGA_2000 %>% count(PAT_NO, ADM_DATE, ADM_HR) %>%
  arrange(desc(n)) %>%
  slice(1)

```
Yep, no duplicates anymore. **PAT_NO-ADM_DATE-ADM_HR** is the composite key.


It appears since we don't have a patient demographics table that we're supposed to get all our data from here. Which variables are these and are they consistent across the table?
```{r}
OMEGA_2000 %>%
  select(PAT_NO,
         AGE, 
         AGE_DAYS, 
         BIRTH_WGHT, # note this variable listed as BIRTH_WEIGHT in the code book
         RACE,
         SEX,
         RELIGION,
         DISTANCE,
         ZIP) %>%
  group_by(PAT_NO,
           AGE, 
           AGE_DAYS, 
           BIRTH_WGHT, 
           RACE,
           SEX,
           RELIGION,
           DISTANCE,
           ZIP) %>%
  mutate(totes_count = n()) %>%
  filter(totes_count > 1) %>%
  arrange(PAT_NO)
```
We have duplicates across all these variables. Let's just remove all duplicates and save in a new dataset for demographic characteristics
```{r}
OMEGA_2000_demo <- OMEGA_2000 %>%
  select(PAT_NO,
         AGE, 
         AGE_DAYS, 
         BIRTH_WGHT, # note this variable listed as BIRTH_WEIGHT in the code book
         RACE,
         SEX,
         RELIGION,
         DISTANCE,
         ZIP) %>%
  unique() 
```


## OME_PX_2000 | Table of all patient operative data

IS PAT_NO-ADM_DATE-OCCURRENCE the primary key?
```{r}
OME_PX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n)) %>%
  slice(1)
```
We have duplicates. Let's examine what's going on within the duplicates.

```{r}
OME_PX_2000_test_1 <- OME_PX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  mutate(totes = n()) %>%
  ungroup() %>%
  filter(totes > 1) %>%
  arrange(PAT_NO)
```

Maybe we need to add the procedure code.

```{r}
OME_PX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE, OP_PROC_9) %>%
  count() %>%
  filter(n > 1) %>%
  arrange(desc(n))
```
Still duplicates. Are these genuinely duplicated rows?

```{r}
OME_PX_2000 %>%
  summarize(n())

OME_PX_2000 %>%
  distinct() %>%
  summarize(n())
```

Ah, there genuinely are duplicates across all rows. Let's remove them.

```{r}
OME_PX_2000 <- OME_PX_2000 %>%
  distinct()
```

Re-check for primary key from PAT_NO-ADM_DATE_OCCURENCE
```{r}
OME_PX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  count() %>%
  arrange(desc(n))
```
Ok, we have our primary key: **PAT_NO-ADM_DATE-OCCURRENCE**

An occurrence is a time during the entire admission that services were provided.

The TOT_NO_OPS column isn't really necessary. It's the max # of occurrences per PAT_NO-ADM_DATE


## OME_DX_00 | Table of all patient diagnostic data

What's the primary key?
```{r}
OME_DX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n)) %>%
  slice(1)
```
We have duplicates across PAT_NO-ADM_DATE-OCCURRENCE. Let's check for genuine row duplication like before.

```{r}
OME_DX_2000 %>%
  summarize(n())

OME_DX_2000 %>%
  distinct() %>%
  summarize(n())
```

Yep, lots of duplicates. Let's remove and re-check the primary key.

```{r}
OME_DX_2000 <- OME_DX_2000 %>% distinct()
```

```{r}
OME_DX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n))%>%
  slice(1)
```

Still duplicates! What else is causing it?

```{r}
OME_DX_2000_test_1 <- OME_DX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  mutate(totes_count = n()) %>%
  ungroup() %>%
  group_by(PAT_NO, ADM_DATE) %>%
  mutate(any_dupes = max(totes_count)) %>%
  ungroup() %>%
  filter(any_dupes > 1) %>%
  arrange(PAT_NO, ADM_DATE, OCCURRENCE)
```
It looks like there can be multiple diagnoses per occurrence.

Let's check counts for a combined key of PAT_NO-ADM_DATE-OCCURRENCE-DIAGNOSIS
```{r}
OME_DX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE, DIAGNOSIS) %>%
  arrange(desc(n))%>%
  slice(1)
```
Still duplicates!! What's going on?

Let's see which cases are throwing a problem
```{r}
OME_DX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE, DIAGNOSIS) %>%
  mutate(totes = n()) %>%
  ungroup() %>%
  filter(totes > 1) %>%
  arrange(PAT_NO, ADM_DATE, OCCURRENCE, DIAGNOSIS)
```

Patient # 28940773 has two rows where everything else is the same except for discharge date. I have no idea what's going on.
Question for DAVID.


# Referential Integrity

```{r}
OMEGA_2000_pat_count <- OMEGA_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

```{r}
OME_DX_2000_pat_count <- OME_DX_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

```{r}
OME_PX_2000_pat_count <- OME_PX_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

```{r}
CRIT_2000_pat_count <- CRIT_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

There are:
`r OMEGA_2000_pat_count` OMEGA_2000 unique patients
`r OME_DX_2000_pat_count` OME_DX_2000 unique patients
`r OME_PX_2000_pat_count` OME_PX_2000 unique patients
`r CRIT_2000_pat_count` CRIT_2000_pat_counts

## Why is the a diagnosis variable in OMEGA_2000 and also in OME_DX_2000?

How do you link the two tables when the primary key of OMEGA_2000 relies on the ADM_HR variable that's not in OME_DX_2000?



# ```{r}
# 
# OMEGA_2000 %>% outer_join(OME_DX_2000, by =c("PAT_NO" = "PAT_NO", 
#                                              "ADM_DATE" = "ADM_DATE",
#                                              "")
# 
# ```

## Do both OMEGA_2000 ADM_SVC and OMEGA2000 DIS_SVC point to the SERVICE table?

# Project questions

## Assessment of the Overall Patient Population and Current Facility Services (20%)



### Top disease entities treated by each group (Cardiac and Orthopedic) as well as the number of patients in each of these disease entities



### Number of patients admitted for Cardiac and Orthopedic services.



### Demographics of these patient groups, including geographic location



### Relevant admission and discharge (disposition) data for these patient groups



### LOS (length of stay) information for these patient groups



### Qualitative assessment of the importance of Cardiac and Orthopedic services



## Intensity/Utilization Indicators (20%)
The issues in this area include:

### What are some key data clues that suggest that the EHS Executive Committee should or should not accept risk via contracts for Cardiac and Orthopedic services?

### What is the current status of the quality of care received by patients admitted via the ER?

### What is the intensity of resource utilization for these services?
A suggested (non-exhaustive) list of evaluation criteria in this area is as follows:

### Reputation of Omega: How far do these patients travel to receive Cardiac and Orthopedic services at Omega compared to other services?

### Death rates

### Percent of encounters where service provided was teaching-related

### Intensity/utilization figures relative to national benchmarks per service

### Physician/surgeon information (e.g., admitting physician, specialty type, operating surgeon)
Percent of visits requiring an external review

### ICU/CCU information (e.g., was a stay in the ICU required; percent distribution of patients by LOS in CCUs; type of CCU where care was received, ALOS - average length of stay - in each unit - type)

### Principal operations performed on each group (e.g., the number of patients having
operations, the average number of operations performed, major types of operations





III.
Financial Viability (20%)
### A suggested (non-exhaustive) list of financial measures includes:

### Profitability of patients from Cardiac and Orthopedic services

### LOS outlier rates

### Payer mix

### Rates of emergent vs. elective care

### Probable future demand given Omega's patient demographics for Cardiac and Orthopedic services

### Potential financial hazards as reimbursement switches from fee-for-service, to case (DRG)-based, to capitation, to capitation with prospective risk -sharing arrangements


