---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---


```{r warning=FALSE, echo=FALSE}

###################################################################
# Load Libraries 
###################################################################
library(tidyverse)

###################################################################
# Load Tables that are stored as CSVs on Github
###################################################################

# ICD-9 Diagnostic codes
#Diganostics 
nineDX11 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/9DX11.csv")

# ICD-9-CM Operative Codes
#Operation 
nine0P11 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/9OP11.csv?token=AOUMBHAGKGSZG2TSGOG2CP3AM2FTQ")

# Patient ICU stays
CRIT_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/CRIT_2000.csv?token=AOUMBHADQBHKXUUKB7V3Q4TAM2GDY") %>%
  select(-PAT_NO_TAG)

# Disposition codes
DISPOSTN <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/DISPOSTN.csv?token=AOUMBHE4FATYKQPJTDSMMF3AM2GE4")

# DRGs
DRG <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/DRG.csv?token=AOUMBHFBTBYE3Q4TL4KTRRDAM2GF4")

# Inpatient encounters
OMEGA_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/OMEGA_2000.csv?token=AOUMBHA2FQ7AWMLVZE5FCQ3AM2GHG")

# Patient diagnostic data
OME_DX_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/OME_DX_2000.csv?token=AOUMBHFOYLTZQZQBGMTEAILAM2GIU")

# Patient operative data
OME_PX_2000 <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/OME_PX_2000.csv?token=AOUMBHA66FZEP7AHOSTBOKDAM2GKY")

# Organ donations
ORGAN <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/ORGAN.csv?token=AOUMBHFMVUHRBMG6KI3IDRDAM2GMC")


##Patients payment method, total, and cost
PAYEX <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/PAYEX.csv?token=AOUMBHDRDOXPNZORSPDOJSLAM2GNC")

# Services codes
SERVICE <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/SERVICE.csv?token=AOUMBHBANABWFQP7P3GVNALAM2GOC")

# Admission Source
SOURCE <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/SOURCE.csv?token=AOUMBHHIEF5IJH3KPSV45STAM2GPQ")

# Tissues
TISSUE <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/TISSUE.csv?token=AOUMBHFMOZUROW2XOBCCVQTAM2GQM")

# Yes/No designation
YES_NO <- read_csv("https://raw.githubusercontent.com/PositiveSumData/HMP_553/main/YES_NO.csv?token=AOUMBHA2GSK3P6IFX4HAYJDAM2GTC")

```

# Examining Referential Integrity

## CRIT_2000 | Patient data table regarding ICU stays

No primary key. Is PAT_NO-ADM_DATE the composite key?
```{r}
CRIT_2000 %>%
  count(PAT_NO, ADM_DATE) %>%
  arrange(desc(n)) %>%
  slice(1)

duplicated(CRIT_2000$PAT_NO) #NOT PK 
duplicated(CRIT_2000$ADM_DATE) #NOT PK

```
Nope, a single PAT_NO-ADM_DATE combo appears six times.

Let's try adding in ICU.
```{r}
CRIT_2000 %>% count(PAT_NO, ADM_DATE, ICU) %>%
  arrange(desc(n))
```
Nope, didn't work either. Let's try PAT_NO-ADM_DATE-OCCURRENCE

```{r}
CRIT_2000 %>% count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n)) %>%
  slice(1)
```
That worked. Our **composite key is PAT_NO-ADM_DATE-OCCURRENCE**.




And what's the difference between all the ICU categories?
```{r}
CRIT_2000 %>% count(ICU)
```
Google says MICU is "medical intensive care unit". SICU is "surgical intensive care unit". I can't seem to figure out what CCU and SPEC_NSRY stand for.

Intensive care units were grouped into 4 types: medical, including coronary care; surgical, including trauma and cardiovascular; neonatal and pediatric; and medical-surgical.
CCU is " more specialized kind of critical care unit, known as a cardiac or coronary care unit"
SPEC_NSRY maybe neonatal and pediatric or critical nursing 


 
What's the difference between ICU_LOS and TOT_CCUS? Is "critical care" different from "intensive care"? Both lengths of stay added together are usually far less than the total stay (discharge - admitted). 

ICU_LOS counts for the 4 different units 
TOT_CCUS the total stay for critical care unit 

A = CRIT_2000$ADM_DATE
D = CRIT_2000$DIS_DATE
# Get the total stay (d-a)
#stay in icu or ccu is after an opertaion, makes sense that is lower 



## DRG

What's the primary key in the DRG table?
```{r}
DRG %>% count(DRG) %>%
  arrange(desc(n)) %>%
  slice(1)
duplicated(DRG$MAJ_DX_CAT)
```
It looks like **DRG is the DRG Primary key**.

What is the WEIGHT and ARITH_LOS and MEAN_LOS and CUTOFF and how is reimbursement calculated? 

```{r}
#Weight =  relative cost weight specific for each DR
#Not sure about the other veribles 
```

Are we supposed to use these in calculating reimbursement by DRG?
#Types of DRGs
DRG %>% count(PARTITION)

#DRG %>% group_by(PARTITION) %>% filter((MEDICAL, NEITHER, PROCEDURAL)) %>% ungroup
```{r}
#analyze the WEIGHT:
fivenum(DRG$WEIGHT)
W = DRG$WEIGHT
#ARITH_LOS, 
fivenum(DRG$ARITH_LOS)
#MEAN_LOS, 
fivenum(DRG$MEAN_LOS)
#CUTOFF 
fivenum(DRG$CUTOFF)
```

no
we need to figure out if there are any outliers in each catagory 

yes 
1. DRG-based payment systems are often referred to as “case-based” or “case-mix-based”

2. DRG payment is based on the care given to and resources used by a "typical" patient within the group. 


## nineDX11

What is the primary key in nineDX11?
```{r}
nineDX11 %>% count(DIAG) %>%
  arrange(desc(n)) %>%
  slice(1)
```
It appears there can be up to two rows for each DIAG in the table. Why?
```{r}
nineDX11 %>%
  group_by(DIAG) %>%
  mutate(totes_diags = n()) %>%
  ungroup() %>%
  filter(totes_diags > 1) %>%
  arrange(DIAG)
```
It seems the rows are complete duplicates, but let's check that's true if we do a group_by for all columns. 

```{r}
nineDX11 %>%
  group_by(DIAG) %>%
  mutate(totes_diags = n()) %>%
  ungroup() %>%
  filter(totes_diags > 1) %>%
  group_by(DIAG, DIAG_DESC, DIAG_C, DIAG_5) %>%
  count() %>%
  arrange()
```
It seems indeed these rows are completely duplicated, because none of them had differences across any of the variables. 

Safe to filter out the duplicate values I think?

```{r}
nineDX11 <- nineDX11 %>%
  distinct()
```
**DIAG is the primary key.**

## nineOP11 | Descriptor Table for ICD-9-CM Operative Codes 

What's the primary key for nineOP11?

```{r}
nine0P11 %>%
  count(OP) %>%
  arrange(desc(n)) %>%
  slice(1)
```
Seems like **OP is the primary key for nineOP11**. All fine here. Could probably use OP_C as well.


## DISPOSTN

Is DISPOSTN the primary key of DISPOSTN?
```{r}
DISPOSTN %>% count(DISPOSTN) %>%
  arrange(desc(n)) %>%
  slice(1)
```
Yep, all fine here. 


## PAYEX

```{r}
There's no table description for PAYEX in the documentation. Does that mean it's safe to assume we don't need to use it?
6 TYPES OF PAYMENTS WITH LOS AND TOT FOR EACH PA 
not sure what to check there 

PAYEX %>% group_by(PRIM_PAY) 
```




## ORGAN

Small table, looks fine. **ORGAN_DON** is the primary key.


## SERVICE

Small table, looks fine. **SERVICE** is the primary key.

## SOURCE 

Small table, looks fine. **ADM_SOURCE** is the primary key.


## TISSUE

Small table, looks fine. **TISSUE** is the primary key.


## YES_NO

Small table, looks fine. **YES_NO** is the primary key. But where does this link to other tables?


## OMEGA_2000 | Data for each inpatient encounter during 2000

What's the primary key in the OMEGA_2000 table?

Start with PAT_NO & ADM_DATE & ADM_HR
```{r}
OMEGA_2000 %>% count(PAT_NO, ADM_DATE, ADM_HR) %>%
  arrange(desc(n)) %>% 
  slice(1) 
```
Nope, we have duplicates of that pairing.

Are there genuinely duplicated rows in the table?

```{r}
OMEGA_2000 %>%
  summarize(n())
OMEGA_2000 %>%
  distinct() %>%
  summarize(n())
```

Yes, it seems there are. Let's remove duplicates.
```{r}
OMEGA_2000 <- OMEGA_2000 %>%
  distinct()
```

Let's check if PAT_NO-ADM_DATE-ADM_HR is now a primary key now that overall duplicates removed 

```{r}
OMEGA_2000 %>% count(PAT_NO, ADM_DATE, DIS_DATE) %>%
  arrange(desc(n)) %>%
  slice(1)

#no need for three composites 
#can eliminate 
#want to do this 
```
Yep, no duplicates anymore. **PAT_NO-ADM_DATE-ADM_HR** is the composite key.


It appears since we don't have a patient demographics table that we're supposed to get all our data from here. Which variables are these and are they consistent across the table?
```{r}
OMEGA_2000 %>%
  select(PAT_NO,
         AGE, 
         AGE_DAYS, 
         BIRTH_WGHT, # note this variable listed as BIRTH_WEIGHT in the code book
         RACE,
         SEX,
         RELIGION,
         DISTANCE,
         ZIP) %>%
  group_by(PAT_NO,
           AGE, 
           AGE_DAYS, 
           BIRTH_WGHT, 
           RACE,
           SEX,
           RELIGION,
           DISTANCE,
           ZIP) %>%
  mutate(totes_count = n()) %>%
  filter(totes_count > 1) %>%
  arrange(PAT_NO)
```
We have duplicates across all these variables. Let's just remove all duplicates and save in a new dataset for demographic characteristics
```{r}
OMEGA_2000_demo <- OMEGA_2000 %>%
  select(PAT_NO,
         AGE, 
         AGE_DAYS, 
         BIRTH_WGHT, # note this variable listed as BIRTH_WEIGHT in the code book
         RACE,
         SEX,
         RELIGION,
         DISTANCE,
         ZIP) %>%
  unique() 
```


## OME_PX_2000 | Table of all patient operative data

IS PAT_NO-ADM_DATE-OCCURRENCE the primary key?
```{r}

OME_PX_2000 <- OME_PX_2000 %>% distinct()

OME_PX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n)) %>%
  slice(1)
```
We have duplicates. Let's examine what's going on within the duplicates.

```{r}
OME_PX_2000_test_1 <- OME_PX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  mutate(totes = n()) %>%
  ungroup() %>%
  filter(totes > 1) %>%
  arrange(PAT_NO)
```

Maybe we need to add the procedure code.

```{r}
OME_PX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE, OP_PROC_9) %>%
  count() %>%
  filter(n > 1) %>%
  arrange(desc(n))
```
Still duplicates. Are these genuinely duplicated rows?

```{r}
OME_PX_2000 %>%
  summarize(n())

OME_PX_2000 %>%
  distinct() %>%
  summarize(n())
```

Ah, there genuinely are duplicates across all rows. Let's remove them.

```{r}
OME_PX_2000 <- OME_PX_2000 %>%
  distinct()
```

Re-check for primary key from PAT_NO-ADM_DATE_OCCURENCE
```{r}
OME_PX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  count() %>%
  arrange(desc(n))
```
Ok, we have our primary key: **PAT_NO-ADM_DATE-OCCURRENCE**

An occurrence is a time during the entire admission that services were provided.

The TOT_NO_OPS column isn't really necessary. It's the max # of occurrences per PAT_NO-ADM_DATE


## OME_DX_00 | Table of all patient diagnostic data

What's the primary key?
```{r}
OME_DX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n)) %>%
  slice(1)
```
We have duplicates across PAT_NO-ADM_DATE-OCCURRENCE. Let's check for genuine row duplication like before.

```{r}
OME_DX_2000 %>%
  summarize(n())

OME_DX_2000 %>%
  distinct() %>%
  summarize(n())
```

Yep, lots of duplicates. Let's remove and re-check the primary key.

```{r}
OME_DX_2000 <- OME_DX_2000 %>% distinct()
```

```{r}
OME_DX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  arrange(desc(n))%>%
  slice(1)
```

Still duplicates! What else is causing it?

```{r}
OME_DX_2000_test_1 <- OME_DX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE) %>%
  mutate(totes_count = n()) %>%
  ungroup() %>%
  group_by(PAT_NO, ADM_DATE) %>%
  mutate(any_dupes = max(totes_count)) %>%
  ungroup() %>%
  filter(any_dupes > 1) %>%
  arrange(PAT_NO, ADM_DATE, OCCURRENCE)
```
It looks like there can be multiple diagnoses per occurrence.

Let's check counts for a combined key of PAT_NO-ADM_DATE-OCCURRENCE-DIAGNOSIS
```{r}
OME_DX_2000 %>%
  count(PAT_NO, ADM_DATE, OCCURRENCE, DIAGNOSIS) %>%
  arrange(desc(n))%>%
  slice(1)
```
Still duplicates!! What's going on?

Let's see which cases are throwing a problem
```{r}
OME_DX_2000 %>%
  group_by(PAT_NO, ADM_DATE, OCCURRENCE, DIAGNOSIS) %>%
  mutate(totes = n()) %>%
  ungroup() %>%
  filter(totes > 1) %>%
  arrange(PAT_NO, ADM_DATE, OCCURRENCE, DIAGNOSIS)
```

Patient # 28940773 has two rows where everything else is the same except for discharge date. I have no idea what's going on.

we might delete
Question for DAVID.


# Referential Integrity

```{r}
OMEGA_2000_pat_count <- OMEGA_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

```{r}
OME_DX_2000_pat_count <- OME_DX_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

```{r}
OME_PX_2000_pat_count <- OME_PX_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

```{r}
CRIT_2000_pat_count <- CRIT_2000 %>% distinct(PAT_NO) %>%
  count() %>% pull()
```

There are:
`r OMEGA_2000_pat_count` OMEGA_2000 unique patients
`r OME_DX_2000_pat_count` OME_DX_2000 unique patients
`r OME_PX_2000_pat_count` OME_PX_2000 unique patients
`r CRIT_2000_pat_count` CRIT_2000_pat_counts

## Why is the a diagnosis variable in OMEGA_2000 and also in OME_DX_2000?

How do you link the two tables when the primary key of OMEGA_2000 relies on the ADM_HR variable that's not in OME_DX_2000?



```{r}


inner_join(OMEGA_2000, OME_DX_2000, by = c("PAT_NO" = "PAT_NO", 
                                              "ADM_DATE" = "ADM_DATE"))

```

Do both OMEGA_2000 ADM_SVC and OMEGA2000 DIS_SVC point to the SERVICE table?
```{r}

OMEGA_2000 %>%
  count(ADM_SVC)
  
  OMEGA_2000 %>%
  count(DIS_SVC)
  
  yes 
```
  

# Project questions

##ICU/CCU information (e.g., was a stay in the ICU required; percent distribution of patients by LOS in CCUs; type of CCU where care was received, ALOS -average length of stay- in each unit-type)

```{r}
#required 


#percent distribution of patients by LOS in CCUs


#CCU TYPE & LOS each type
types = CRIT_2000 %>% count(ICU)
ggplot(data=CRIT_2000, mapping=aes(x=CRIT_2000$ICU, y=CRIT_2000$ICU_LOS) + geom_bar(stat='identity')



```



##Principal operations performed on each group (e.g., the number of patients having operations, the average number of operations performed, major types of operations performed)
```{r}
#the number of patients having operations
OME_PX_2000 %>% count(PAT_NO)

#the average number of operations performed

#major types of operations performed

```

##Reputation of Omega: How far do these patients travel to receive Cardiac and Orthopedic services at Omega compared to other services?
```{r}
# How far do these patients travel to receive Cardiac and Orthopedic services w'Omega compared vs others
```

##Death rates

```{r}
```



III.
Financial Viability (20%)
### A suggested (non-exhaustive) list of financial measures includes:

### Profitability of patients from Cardiac and Orthopedic services

### LOS outlier rates

### Payer mix

### Rates of emergent vs. elective care

### Probable future demand given Omega's patient demographics for Cardiac and Orthopedic services

### Potential financial hazards as reimbursement switches from fee-for-service, to case (DRG)-based, to capitation, to capitation with prospective risk -sharing arrangements


